Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-27 paper-new/discussion.tex
discussion revisions

\subsection{Related Work}
\max{review what I wrote, and discuss Jeremy's work}
% Discuss Joey Eremondi's thesis on gradual dependent types
Eremondi \cite{Eremondi_2023} uses guarded type theory to
define a syntactic model for a gradually-typed source
language with dependent types. By working in guarded type theory, they are
able to give an exact semantics to non-terminating computations in a language
which is logically consistent, allowing for metatheoretic results about the
source language to be proven.
%
Similarly to our approach, they define a guarded lift monad to model potentially-
nonterminating computations and use guarded recursion to model the dynamic type.
However, they do not give a denotational semantics to term precision and it is unclear
how to prove the gradual guarantee in this setting.
The work includes a formalization of the syntactic model in Guarded Cubical Agda.

@@

\subsection{Mechanization}
\max{review}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper,
we are mechanizing our results and developing a reusable framework
for proving graduality in Guarded Cubical Agda \cite{veltri-vezzosi2020}.

As of this writing, the work is in progress, but we have
constructed most parts of the concrete model discussed in
Section \ref{sec:concrete-model}. For instance, we have defined
types for predomains, error domains, and their morphisms and relations,
and we used the guarded features to define the guarded lift + error monad
and the dynamic type.

@@

Lastly, we would like to formalize the adequacy result discussed in
Section \ref{sec:adequacy}. This will involve adding axioms about
clock quantification as well as about the clock-irrelevance of
booleans and natural numbers, since as of this writing these
axioms are not built-in to Guarded Cubical Agda. As an experiment, we have 
formalized some basic results involving clocks as part of our development.

@@

\subsection{Comparison to Explicit Step-Indexing}
\max{check/revise/delete if it's not important}
Working internally to guarded type theory reduces the overhead of needing to
carry around the step-indices in the proofs as is required when using explicit
step-indexing. Additionally, the logical relations constructed to prove
graduality suffer from technical complications of requiring two separate
expression relations, one that counts steps on the left and the other on the right.
There is no analogue of this in our approach.
However, the price we pay for this simplicity is that it is unclear
whether using our approach one can obtain an extensional model that models transitivity
of term precision, which is attainable using the traditional step-indexing approach.

\subsection{Synthetic Ordering}
\max{revise (fix citations) or remove if it's no longer relevant}
A key to managing the complexity of our concrete construction is in
using a \emph{synthetic} approach to step-indexing rather than working
analytically with presheaves. This has helped immensely in our ongoing
mechanization in cubical Agda as it sidesteps the need to formalize
these constructions internally. 
%
However, there are other aspects of the model, the bisimilarity and
the monotonicity, which are treated analytically and are similarly
tedious.
%
It may be possible to utilize further synthetic techniques to reduce
this burden as well, and have all type intrinsically carry a notion of
bisimilarity and ordering relation, and all constructions to
automatically preserve them.
%
A synthetic approach to ordering is common in (non-guarded) synthetic
domain theory and has also been used for synthetic reasoning for cost
models \cite{synthetic-domain-theory,decalf}.

\subsection{Future Work}
\max{review and change if needed}
We would like to apply our approach to give a denotational semantics to gradually-typed
languages with algebraic effects, building on prior work on gradual typing for effect handlers
\cite{greff}. In particular, that work proves graduality via a complicated step-indexed logical relation,
and we hope to prove their results by building a denotational model for GrEff.
This would serve as a step towards applying our techniques to prove graduality for languages
with other advanced features.

The extensional model we construct differs from the usual notion of extensional
model considered in prior work on gradual typing in that it lacks horizontal composition of squares.
We would like to clarify the relationship between our notion of model and prior extensional models,
with the aim of determining whether our approach could allow for the construction of such a model.

>>>>

\subsection{Related Work}
% Discuss Joey Eremondi's thesis on gradual dependent types
Eremondi \cite{Eremondi_2023} uses guarded type theory to
define a syntactic model for a gradually-typed source
language with dependent types. By working in guarded type theory, they are
able to give an exact semantics to non-terminating computations in a language
which is logically consistent, allowing for metatheoretic results about the
source language to be proven.
%
Similarly to our approach, they define a guarded lift monad to model potentially-
nonterminating computations and use guarded recursion to model the dynamic type.
However, they do not give a denotational semantics to term precision and it is unclear
how to prove the gradual guarantee in this setting.
The work includes a formalization of the syntactic model in Guarded Cubical Agda.

@@

\subsection{Mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper,
we are mechanizing our results and developing a reusable framework
for proving graduality in Guarded Cubical Agda \cite{veltri-vezzosi2020}.
%
As of this writing, the work is in progress, but we have
constructed most parts of the concrete model discussed in
Section \ref{sec:concrete-model}. For instance, we have defined
types for predomains, error domains, and their morphisms and relations,
and we used the guarded features to define the guarded lift + error monad
and the dynamic type.

@@

Lastly, we plan to formalize the adequacy result discussed in
Section \ref{sec:adequacy}. This will involve adding axioms about
clock quantification as well as about the clock-irrelevance of
booleans and natural numbers, since as of this writing these
axioms are not built-in to Guarded Cubical Agda. As an experiment, we have 
formalized some basic results involving clocks as part of our development.

@@

\subsection{Comparison to Explicit Step-Indexing}

Working internally to guarded type theory reduces the overhead of
needing to carry around the step-indices in the proofs as is required
when using explicit step-indexing. Additionally, the logical relations
constructed to prove graduality in prior work \cite{newahmed,gtt,greff} suffer
from technical complications of requiring two separate expression
relations, one that counts steps on the left and the other on the
right, and there is no analogue of this in our approach. However,
using two expression relations allows some but not all transitivite
reasoning of term precision to be recovered. In the future we aim to
explore if this approach is feasible in guarded semantics.

\subsection{Synthetic Ordering}
A key to managing the complexity of our concrete construction is in
using a \emph{synthetic} approach to step-indexing rather than working
analytically with presheaves. This has helped immensely in our ongoing
mechanization in cubical Agda as it sidesteps the need to formalize
these constructions internally. 
%
However, there are other aspects of the model, the bisimilarity and
the monotonicity, which are treated analytically and are similarly
tedious.
%
It may be possible to utilize further synthetic techniques to reduce
this burden as well, and have all type intrinsically carry a notion of
bisimilarity and ordering relation, and all constructions to
automatically preserve them.
%
A synthetic approach to ordering is common in (non-guarded) synthetic
domain theory and has also been used for synthetic reasoning for cost
models \cite{synthetic-domain-theory,decalf}.

\subsection{Future Work}

In the future, we plan to apply our approach to give a denotational
semantics for languages that feature higher-order state or
runtime-extensible dynamic typing \cite{guarded-higherorderstate} as
well as richer type disciplines such as gradual dependent types and
effect systems.

%% to gradually-typed
%% languages with algebraic effects, building on prior work on gradual typing for effect handlers
%% \cite{greff}. In particular, that work proves graduality via a complicated step-indexed logical relation,
%% and we hope to prove their results by building a denotational model for GrEff.
%% This would serve as a step towards applying our techniques to prove graduality for languages
%% with other advanced features.

%% The extensional model we construct differs from the usual notion of extensional
%% model considered in prior work on gradual typing in that it lacks horizontal composition of squares.
%% We would like to clarify the relationship between our notion of model and prior extensional models,
%% with the aim of determining whether our approach could allow for the construction of such a model.
